---
title: "Johns Hopkins University - Data Science - Reproducible Research"
author: "Alvaro Lozano Alonso"
date: "1/7/2021"
output: html_document
---

## Introduction

This file is the Course Project 1 report of the Reproducible Research course, which is part of the Data Science Program of the Johns Hopkins University.

## Code

Setup code.  
```{r setup}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(warning = FALSE)
knitr::opts_chunk$set(message = FALSE)
```

First we load the necessary libraries that are going to be useful to process the data.  
```{r}
library(dplyr)
library(ggplot2)
library(tidyverse)
library(lubridate)
library(lattice)
```

### 1. Code for reading in the dataset and/or processing the data.
```{r}
setwd("~/Documents/Dokumente - MacBook Air/Data Science with R - Johns Hopkins University/Reproducible Research/JHU-Data_Science-Reproducible_Research")
activity <- read.csv("activity.csv", header=TRUE, sep=",")
head(activity)
str(activity)
```

### 2. Histogram of the total number of steps taken each day.
```{r}
activityPerDay <- activity %>% group_by(date) %>% summarise(Steps = sum(steps, na.rm=TRUE))
ggplot(activityPerDay, aes(x=Steps)) + 
  geom_histogram(fill="#69b3a2", color="#e9ecef", alpha=0.9) +
  ylab("Frequency") +
  xlab("Steps") +
  ggtitle("Histogram steps per day")
```
  
### 3. Mean and median number of steps taken each day.
```{r}
stepsPerDayMean <- activity %>% group_by(date) %>% summarise(Mean = mean(steps, na.rm=TRUE))
ggplot(stepsPerDayMean, aes(x=date, y=Mean)) + 
  geom_bar(fill="#69b3a2", stat = "identity") + 
  theme(axis.text.x=element_blank(),
        axis.ticks.x=element_blank()) +
  ggtitle("Mean of steps per 5 minutes interval") +
  ylab("Mean per 5 minutes interval")

stepsPerDayMedian <- activity %>% group_by(date) %>% summarise(Median = median(steps, na.rm=TRUE))
ggplot(stepsPerDayMedian, aes(x=date, y=Median)) + 
  geom_point(fill="#69b3a2", stat = "identity") + 
  theme(axis.text.x=element_blank(),
        axis.ticks.x=element_blank()) +
  ggtitle("Median of steps per 5 minutes interval") +
  ylab("Median per 5 minutes interval")

totalStepsDay <- activity %>% group_by(date) %>% summarise(TotalSteps = sum(steps, na.rm=TRUE))
mymean <- mean(totalStepsDay$TotalSteps)
median <- as.numeric(median(totalStepsDay$TotalSteps))

print(paste("The mean steps per day is: ", mymean))
print(paste("The median steps per day is: ", median))
```
  
### 4. Time series plot of the average number of steps taken.
```{r}
stepsPerIntervalAverage <- activity %>% group_by(interval) %>% summarise(StepsAverage = mean(steps, na.rm=TRUE))
ggplot(stepsPerIntervalAverage, aes(x=interval, y=StepsAverage)) +
  geom_line(color="#69b3a2", alpha=0.9) + 
  xlab("Interval") +
  ggtitle("Time series of the average number of steps")
```
  
### 5. The 5-minute interval that, on average, contains the maximum number of steps.
```{r}
maxSteps <- max(stepsPerIntervalAverage[,2])
interval <- stepsPerIntervalAverage[which(stepsPerIntervalAverage$StepsAverage == maxSteps), 1]
print(paste("The maximum number of steps per day is: ", maxSteps, "And the the interval is: ", as.numeric(interval)))
```
### 6. Code to describe and show a strategy for imputing missing data.
```{r}
activity.noNA <- activity
print(paste("The total number of rows with NA values is: ", sum(is.na(activity.noNA$steps))))
for (i in 1:length(activity.noNA$steps)) {
  if (is.na(activity.noNA[i, 1])) {
     activity.noNA[i, 1] <- stepsPerIntervalAverage[match(activity.noNA[i, 3], stepsPerIntervalAverage$interval), 
                                                    2]
  }
}
```

### 7. Histogram of the total number of steps taken each day after missing values are imputed.
```{r}
activityPerDay.noNA <- activity.noNA %>% group_by(date) %>% summarise(Steps = sum(steps, na.rm=TRUE))
ggplot(activityPerDay.noNA, aes(x=Steps)) + 
  geom_histogram(fill="#69b3a2", color="#e9ecef", alpha=0.9) +
  ylab("Frequency") +
  xlab("Steps") +
  ggtitle("Histogram steps per day")

stepsPerDayMean.noNA <- activity.noNA %>% group_by(date) %>% summarise(Mean = mean(steps, na.rm=TRUE))
ggplot(stepsPerDayMean.noNA, aes(x=date, y=Mean)) + 
  geom_bar(fill="#69b3a2", stat = "identity") + 
  theme(axis.text.x=element_blank(),
        axis.ticks.x=element_blank()) +
  ggtitle("Mean of steps per 5 minutes interval") +
  ylab("Mean per 5 minutes interval")

stepsPerDayMedian.noNA <- activity.noNA %>% group_by(date) %>% summarise(Median = median(steps, na.rm=TRUE))
ggplot(stepsPerDayMedian.noNA, aes(x=date, y=Median)) + 
  geom_point(fill="#69b3a2", stat = "identity") + 
  theme(axis.text.x=element_blank(),
        axis.ticks.x=element_blank()) +
  ggtitle("Median of steps per 5 minutes interval") +
  ylab("Median per 5 minutes interval")

totalStepsDay.noNA <- activity.noNA %>% group_by(date) %>% summarise(TotalSteps = sum(steps, na.rm=TRUE))
mean.noNA <- mean(totalStepsDay.noNA$TotalSteps)
median.noNA <- as.numeric(median(totalStepsDay.noNA$TotalSteps))

print(paste("The mean steps per day is: ", mean.noNA))
print(paste("The median steps per day is: ", median.noNA))
```
```{r}
ggplot(stepsPerDayMean, aes(x=date, y=Mean)) + 
  geom_bar(fill="#69b3a2", stat = "identity") + 
  theme(axis.text.x=element_blank(),
        axis.ticks.x=element_blank())

ggplot(stepsPerDayMean.noNA, aes(x=date, y=Mean)) + 
  geom_bar(fill="#69b3a2", stat = "identity") + 
  theme(axis.text.x=element_blank(),
        axis.ticks.x=element_blank())
```
   
### 8. Panel plot comparing the average number of steps taken per 5-minute interval across weekdays and weekends.
```{r}
activity.noNA$date <- ymd(activity.noNA$date)
activity.noNA$weekday <- weekdays(activity.noNA$date)
activity.noNA$daytype <- ifelse(activity.noNA$weekday %in% c("Saturday", "Sunday"), "Weekend", "Weekday")
activity.noNA$daytype <- as.factor(activity.noNA$daytype)
```
  

```{r}
stepsPerIntervalAverage.noNA.weekdays <- activity.noNA %>% filter(daytype=="Weekday") %>% group_by(interval) %>% summarise(StepsAverage = mean(steps, na.rm=TRUE))

stepsPerIntervalAverage.noNA.weekends <- activity.noNA %>% filter(daytype=="Weekend") %>% group_by(interval) %>% summarise(StepsAverage = mean(steps, na.rm=TRUE))
```

```{r}
ggplot(stepsPerIntervalAverage.noNA.weekdays, aes(x=interval, y=StepsAverage)) +
  geom_line(color="#69b3a2", alpha=0.9) + 
  xlab("Interval") +
  ggtitle("Time series of the average number of steps")
```

```{r}
ggplot(stepsPerIntervalAverage.noNA.weekends, aes(x=interval, y=StepsAverage)) +
  geom_line(color="#69b3a2", alpha=0.9) + 
  xlab("Interval") +
  ggtitle("Time series of the average number of steps")
```
```{r}
plotdata <- aggregate(steps ~ interval + daytype, activity.noNA, mean)
xyplot(steps ~ interval | factor(daytype), data=plotdata, aspect=1/3, type="l")
```










